\section{Related Work}  
\label{sec:related_work}

\textcolor{red}{sleng med noe om fashion-mnist og k-mnist her??}

\subsection{Hardware Accelerators for Tsetlin Machines}
\label{sec:related_hw}

The first realized HW implementation of the TM was introduced in 2020. It was implemented both on 65 nm ASIC as well as on an Altera Stratix IV FPGA \textcolor{red}{CITE WHEELDON 2020}. The design supported on-device learning by utilizing parallel clause updates and stochastic feedback, by using lightweight pseudo-random number generators. It demonstrated competitive energy efficiency, outperforming a Convolutional Binary Neural Network by $2.5\times$ in energy per operation. However, the accuracy fell short of the software equivalent of the TM. This difference was attributed to differences in random number generator precision.

Extending this to spatial data, the first hardware accelerator for the Convolutional TM (CTM) was presented in 2023 \textcolor{red}{CITE TUNHEIM 2023}. The design targeted binary image classification and supported both inference and on-device learning, achieving throughput 13.3 times higher than a software implementation running on a 2.7 GHz CPU. The work also identified the quality of the random number generator as a critical factor in achieving accuracy comparable to software, finding a 16-bit LFSR to be the optimal configuration. 

More recently, two hardware accelerators for the Convolutional Coalesced TM (ConvCoTM) were presented in 2025. The first is an FPGA implementation targeting $28 \times 28$ grayscale images across ten output classes, achieving 134k classifications per second and $13.3\,\mu$J energy per classification (EPC) on a Xilinx Zynq ZU7EV while supporting full on-device learning \textcolor{red}{CITE TUNHEIM 2025 FPGA}. The second is a 65 nm ASIC inference-only implementation of the same architecture, achieving 60.3k classifications per second at just 8.6 nJ EPC, one of the lowest reported for TM hardware \textcolor{red}{CITE TUNHEIM 2025 ASIC}. Together, these designs demonstrate how dedicated TM hardware can deliver high energy efficiency and competitive accuracy for edge deployment. 

\subsection{Processor-Oriented TM Acceleration}
\label{sec:related_processor}

Zeng et al.\ propose an optimized software TM inference implementation targeting general-purpose CPUs, evaluated using the gem5 simulator with an ARM processor architecture \textcolor{red}{CITE ZENG 2025}. The key contributions are a bitwise operation based clause evaluation method, an early exit mechanism that terminates clause evaluation as soon as a failing literal group is detected, and a Reorder strategy that reorders literals and TA actions before inference to maximize the likelihood of early exits. Together, these techniques reduce inference time by up to 96.71\% compared to a conventional integer-based baseline, with the most significant gains observed on high-dimensional datasets such as MNIST. Because the approach is purely software-based, it requires no hardware modifications and is directly portable across processor platforms.

\subsection{RISC-V Custom Extensions for Machine Learning}
\label{sec:related_riscv}

Wang et al.\ propose a set of custom RISC-V SIMD instructions targeting the convolution, activation, and pooling operations common in neural network inference \textcolor{red}{CITE WANG 2024}. Rather than relying on a large dedicated accelerator, the approach adds a compact set of specialized instructions that process multiple data values in parallel, achieving meaningful speed and energy efficiency improvements on FPGA-based edge systems while keeping the processor design lightweight and programmable. This work is directly relevant to the approach taken in this thesis, where a custom RISC-V instruction is used to accelerate TM clause evaluation within the NEORV32 processor, targeting a similar balance between programmability and hardware efficiency.

\subsection{Prior Work: Processor-Level Acceleration of Tsetlin Machines
Using Custom RISC-V ISA Extensions}
\label{sec:related_prior}

The present thesis extends the work reported in \textcolor{red}{CITE DEG SJÆL}, which investigated processor-level TM inference acceleration on the same NEORV32 RISC-V soft-core processor deployed on a Xilinx Artix-7 FPGA. That work designed a custom clause evaluation instruction using the CFU, implemented using the R4-type instruction format available in the version of NEORV32 used at the time, and evaluated inference performance on the NoisyXOR dataset.

The results demonstrated a clear speedup for the NoisyXOR problem, where the model is small enough to reside entirely in on-chip DMEM, confirming that the CFU-based approach can meaningfully reduce clause evaluation time when memory access is not the bottleneck. The central question left open by that work is whether the same acceleration benefit scales to larger, more practically relevant problems such as MNIST digit classification, where the model size far exceeds on-chip memory capacity and must be stored in off-chip DDR2 SDRAM. Investigating this question is the primary research objective of the present thesis.