% to be added: introductory paragraph

Hardware accelerators are widely integrated into electronic systems to offload application-specific workloads from general-purpose processors. This is especially relevant for Internet-of-Things (IoT) edge nodes, where compute and memory resources are limited, and battery lifetime is a primary constraint \cite{9985008}. Much of the work focusing on modern machine learning has been targeting convolutional neural networks (CNNs) \cite{Lecun2015436}, which provide state-of-the-art accuracy. However, this performance comes at a price, as neural networks typically require large amounts of data and computational resources \cite{GranmoOle-Christoffer2018TTM-}, which is a difficult match for the resource constrained nature of edge devices. 


% pagraph presenting the Tsetlin Machine, quick introduction of the architecture and how it works, and why it is interesting for hardware acceleration and edge deployment. This will be expanded in the theory chapter, but a quick introduction here will help set the stage for the rest of the thesis.
The Tsetlin Machine (TM) offers a qualitatively different approach to other machine learning models. Introduced by Ole-Christoffer Granmo in 2018 \cite{GranmoOle-Christoffer2018TTM-}, the TM is a logic-based learning algorithm founded on propositional logic and finite-state automata. It constructs propositional formulas (clauses) to perform classification tasks, where each clause is a conjunction of literals (input features and their negations) that vote for or against a particular class. The learning process involves updating the states of Tsetlin Automata based on feedback from the classification results, allowing the model to dynamically refine its decision boundaries. The architecture of the TM enables execution using simple bitwise operations, making it particularly suitable for hardware acceleration and edge deployment \cite{AndersTunheimSvein2025AA86, TunheimSveinAnders2023CTMT, TunheimSveinAnders2025TMIC}. Fixed-function FPGA and ASIC accelerators have demonstrated that TM inference can be executed at high throughput and low energy \cite{AndersTunheimSvein2025AA86, TunheimSveinAnders2023CTMT, TunheimSveinAnders2025TMIC}, but these designs commit to a specific TM architecture at synthesis time and require significant redesign to accommodate new variants or problem configurations.

Processor-oriented acceleration preserves programmability by embedding domain-specific logic inside a general-purpose processor rather than alongside it. The RISC-V Instruction Set Architecture (ISA) is well-suited to this approach, providing reserved opcode space for user-defined instruction extensions that integrate directly into the processor pipeline \cite{WatermanLee2011,riscv_unpriv_isa}. A preceding project thesis \cite{JB2025} demonstrated this on the Noisy XOR (Section \ref{sec:other_datasets}) problem, implementing a custom clause evaluation instruction on the NEORV32 soft-core processor and achieving a $4.42\times$ inference speedup at a cost of only 18 additional LUTs. 

This thesis builds on that work by examining how the approach scales to a realistic workload, specifically MNIST digit classification. Whereas the previous work operated on a model of 480 bytes, this work targets a model of approximately 3.81\,MB, an increase of roughly $8000\times$. A scaling of this magnitude introduces a qualitatively different challenge: the model far exceeds the processor's internal memories, placing memory bandwidth rather than computation at the centre of the performance problem.


\section{UN Sustainability Goals}
This work relates to two of the United Nations Sustainable Development Goals~\cite{UNSDG}. The first is Goal 7, Affordable and Clean Energy. As machine learning inference on edge devices becomes more widespread, the energy it consumes is a growing concern, particularly for battery-powered IoT nodes. The Tsetlin Machine performs inference using only bitwise operations and integer arithmetic, avoiding the multiply-accumulate and floating-point operations that dominate the energy cost of conventional neural networks. Investigating how such a model can be executed efficiently on low-power hardware contributes to reducing the energy demand of on-device intelligence. The second is Goal 9, Industry, Innovation and Infrastructure. This work is built on open and royalty-free foundations, namely the RISC-V instruction set architecture and the open-source NEORV32 soft-core processor, which lower the barrier to developing custom computing hardware. Exploring how application-specific acceleration can be added to an open processor rather than a proprietary platform supports innovation in accessible and adaptable computing
infrastructure.


\input{Chapters/01Introduction/6RelatedWorks.tex}